ODGI: scalable tools for pangenome graphs

### 2021-05-31 manuscript/main.tex
let's try to fit with a Methods section
extended the proposed layout

@@

    \odgi\ has a wide range of commands. Important features are

    \topic{odgi build, view \& validate} pangenome graphs are commonly
    represented by a subset of Graphical Fragment Assembly (GFAv1)
    format\citep{GFAv1}.  \cmd{odgi build} and \cmd{odgi view} allow
    to convert graphs from GFA to ODGI format and vice versa. In the
    variation graph model, paths have to respect the graph's topology:
    this can be verified with \texttt{odgi validate}, to ensure no
    errors in the input or edited graphs.

    \topic{odgi viz, sort, drawn \& layout} by visualizing pangenome
    graphs, we can gain insight into the relationship between
    genomes. \cmd{odgi viz} applies binding and direct rendering to
    a raster image to generate a 1D representation able to scale up to
    gigabase pangenome graphs
    (Figure~1.A\vphantom{\ref{fig:01}}). However, pangenome graphs may
    present complex structures to visualize: \cmd{odgi sort} finds
    the best node order to simplify 1D graph visualization by applying
    several cutting-edge sorting algorithms. \cmd{odgi drawn}
    extends the visualization in 2D, taking in input the layouts built
    by \cmd{odgi layout}.

    \topic{odgi stats} users can retrieve metrics describing the graph properties, such as the number of
    nodes, edges, paths, and graph length. \odgi\ also offers more advanced tools for interrogating the graphs.
    \texttt{odgi bin} summarizes the path information into bins of specified size, enabling a summarized view of
    gigabase scale graphs. \texttt{odgi depth} returns the node depth as defined by query criteria, allowing users to
    detect the complex regions in the graph due to sequences with highly identical repeats. Complex motifs can be also
    detected with \texttt{odgi degree}, which returns the node degree as defined by query criteria. High degree nodes
    are the mirror of misassemblies or problems in the pangenome building, making the tool useful for debugging the
    process as well.

    \odgi\ allows edit operations on the graphs. Cycles in the graph complicate downstream analyses: \texttt{odgi break}
    removes the cycles, reducing the complexity of the graph topology. \texttt{odgi groom} removes spurious inverting
    links by exploring the graph from the orientation supported by most paths. To enable efficient sequence alignment
    against the graph, long nodes can be divided into shorter nodes at a maximum requested size using
    \texttt{odgi chop}. Partial order alignment, which consists of aligning sequences against a directed acyclic graph
    (DAG), is frequently used in pangenome building pipelines, but the current implementations return DAGs with 1-bp
    long nodes; \texttt{odgi unchop} allows joining nodes that can be merged without changing the graph topology, nor
    the embedded sequences, obtaining an equivalent, but more compact, representation of the graph.

    Pangenome graphs can embed multiple chromosomes as separated connected components (inter-chromosomal structural
    variants would join the components into bigger ones). \texttt{odgi explode} separates the connected components in
    different \odgi\ format files, while \texttt{odgi squeeze} allows merging multiple graphs into the same \odgi\ format
    file, preventing node ID collisions. \texttt{odgi extract} allows extracting specific subregions of the graph as
    defined by query criteria, thereby simplifying the downstream analyses and reducing the resources to work only
    with the extracted region.

    In variation graphs the coordinates are provided by the embedded path sequences. Indeed, the node IDs are not
    meant to be stable. \texttt{odgi position} finds, translates, and liftovers graph and path positions between
    different graphs by exploiting their shared path sequences (Figure~1.B\vphantom{\ref{fig:01}}).

    \begin{figure}
        \begin{minipage}[b]{0.48\textwidth}
            \includegraphics[width=\linewidth]{fig/mhc-align.eps}
            \caption{(a) alignment}
            \label{fig:1}
        \end{minipage}

        \begin{minipage}[b]{0.48\textwidth}
            \includegraphics[width=\linewidth]{fig/mhc-pangenome.eps}
            \caption{(b) Visualizing the Major Histocompatibility Complex
                (MHC) \textit{locus} pangenome. A) \texttt{odgi viz} 1D visualization
                of 8 haploid, phased human genome assemblies plus the chm13
                and GRCh38 reference genomes. Graph nodes’ are arranged
                from left to right. The colored bars represent the
                linearized renderings of the embedded paths versus the nodes
                in a binary matrix. The black lines under the paths are the
                links, which represent the graph topology. B)
                Bandage~\citep{26099265} representation of the graph
                topology of a consensus MHC pangenome graph representing
                only variations bigger than 100 bps. Gene labels were
                applied exploiting \texttt{odgi position} to liftover gene
                coordinates from the full graph into the consensus graph.  }

@@

    To work with \odgi\ and VG it is first necessary to convert the graph from GFA to \odgi\ and VG format, respectively.
    Thanks to the \odgi\ format's path implementation, \texttt{odgi build} embeds the paths in parallel, outperforming
    \texttt{vg convert} in the same task. For the same reason, \texttt{odgi extract} results 20 times faster than
    \texttt{vg chunk} in extracting a portion from the full graph.

>>>>

    \section{Methods}

@@

    \odgi\ has a wide range of commands.

    \topic{odgi build, view \& validate} pangenome graphs are commonly
    represented by a subset of Graphical Fragment Assembly (GFAv1)
    format \citep{GFAv1}; \cmd{odgi build} and \cmd{odgi view} allow
    to convert graphs from GFA to \odgi\ format and vice versa. In the
    variation graph model, paths have to respect the graph's topology:
    this can be verified with \cmd{odgi validate}, to ensure no
    errors in the input or edited graphs.

    \topic{odgi viz, sort, drawn \& layout} by visualizing pangenome
    graphs, we can gain insight into the relationship between
    genomes. \cmd{odgi viz} applies binding and direct rendering to
    a raster image to generate a 1D representation able to scale up to
    gigabase pangenome graphs (Figure~1\vphantom{\ref{fig:1}}).
    However, pangenome graphs may present complex structures to
    visualize: \cmd{odgi sort} finds the best node order to simplify
    1D graph visualization by applying several cutting-edge sorting
    algorithms. \cmd{odgi drawn} extends the visualization in 2D,
    taking in input the layouts built by \cmd{odgi layout}.

    \topic{odgi stats, bin, depth \& degree} users can retrieve metrics describing the graph properties, such as the number of
    nodes, edges, paths, and graph length. \odgi\ also offers more advanced tools for interrogating the graphs.
    \cmd{odgi bin} summarizes the path information into bins of specified size, enabling a summarized view of
    gigabase scale graphs. \cmd{odgi depth} returns the node depth as defined by query criteria, allowing users to
    detect the complex regions in the graph due to sequences with highly identical repeats. Complex motifs can be also
    detected with \cmd{odgi degree}, which returns the node degree as defined by query criteria. High degree nodes
    are the mirror of misassemblies or problems in the pangenome building, making the tool useful for debugging the
    process as well.

    \topic{odgi break, groom, chop \& unchop} \odgi\ allows edit operations on the graphs. Cycles in the graph
    complicate downstream analyses: \cmd{odgi break} removes the cycles, reducing the complexity of the graph topology.
    \cmd{odgi groom} removes spurious inverting links by exploring the graph from the orientation supported by most
    paths. To enable efficient sequence alignment against the graph, long nodes can be divided into shorter nodes
    at a maximum requested size using \cmd{odgi chop}. Partial order alignment, which consists of aligning sequences
    against a directed acyclic graph (DAG), is frequently used in pangenome building pipelines, but the current
    implementations return DAGs with 1-bp long nodes; \cmd{odgi unchop} allows joining nodes that can be merged
    without changing the graph topology, nor the embedded sequences, obtaining an equivalent, but more compact,
    representation of the graph.

    \topic{odgi explode, squeeze \& extract} pangenome graphs can embed multiple chromosomes as separated connected
    components (inter-chromosomal structural variants would join the components into bigger ones).
    \cmd{odgi explode} separates the connected components in different \odgi\ format files, while \cmd{odgi squeeze}
    allows merging multiple graphs into the same \odgi\ format file, preventing node ID collisions. \cmd{odgi extract}
    allows extracting specific subregions of the graph as defined by query criteria, thereby simplifying the downstream
    analyses and reducing the resources to work only with the extracted region.

    \topic{odgi position} in variation graphs the coordinates are provided by the embedded path sequences. Indeed, the node
    IDs are not meant to be stable. \cmd{odgi position} finds, translates, and liftovers graph and path positions between
    different graphs by exploiting their shared path sequences (Figure~2\vphantom{\ref{fig:2}}).

    \begin{figure}
        \begin{minipage}[b]{0.48\textwidth}
            \includegraphics[width=\linewidth]{fig/mhc-align}
            \caption{(a) alignment}
            \label{fig:1}
        \end{minipage}

        \begin{minipage}[b]{0.48\textwidth}
            \includegraphics[width=\linewidth]{fig/mhc-pangenome}
            \caption{(b) Visualizing the Major Histocompatibility Complex
                (MHC) \textit{locus} pangenome. A) \cmd{odgi viz} 1D visualization
                of 8 haploid, phased human genome assemblies plus the chm13
                and GRCh38 reference genomes. Graph nodes’ are arranged
                from left to right. The colored bars represent the
                linearized renderings of the embedded paths versus the nodes
                in a binary matrix. The black lines under the paths are the
                links, which represent the graph topology. B)
                Bandage~\citep{26099265} representation of the graph
                topology of a consensus MHC pangenome graph representing
                only variations bigger than 100 bps. Gene labels were
                applied exploiting \cmd{odgi position} to liftover gene
                coordinates from the full graph into the consensus graph.  }

@@

    To work with \odgi\ and VG it is first necessary to convert the graph from GFA to \odgi\ and VG format, respectively.
    Thanks to the \odgi\ format's path implementation, \cmd{odgi build} embeds the paths in parallel, outperforming
    \cmd{vg convert} in the same task. For the same reason, \cmd{odgi extract} results 20 times faster than
    \cmd{vg chunk} in extracting a portion from the full graph.
